# brfssTools — Project Scratchpad

_Last updated: 2026-04-30_

## What this package does (in one sentence)

A standardized framework for pulling, harmonizing, and analyzing BRFSS data
across years and across CDC↔state surveys, with a hand-curated crosswalk
and an interactive editor for maintaining it.

## Architecture at a glance

```{r}
                  Package install                   User config dir
                 (read-only, ships)         (writable, persists across upgrades)
       ┌───────────────────────────────┐   ┌─────────────────────────────────┐
       │ inst/extdata/                 │   │ tools::R_user_dir(              │
       │ ├── cdc_codebook.csv          │   │   "brfssTools", "config")        │
       │ ├── cdc_calculated_vars.csv   │   │ ├── crosswalk.csv               │
       │ ├── cdc_only.csv              │   │ ├── state_codebook.csv          │
       │ ├── taxonomy.csv              │   │ ├── state_only.csv              │
       │ └── (sample OR data,          │   │ ├── pending.csv                 │
       │      copied on init)          │   │ ├── config.yaml                 │
       └───────────────────────────────┘   │ └── transformations/            │
                                           │     ├── race.yaml               │
                                           │     ├── race.R    (autogen)     │
                                           │     └── ... user files ...      │
                                           └─────────────────────────────────┘
```

`brfss_init_state("OR")` copies the package's shipped Oregon sample into
the user's config dir. Other states pass `state_codebook_path = "..."` to
seed an empty crosswalk against their codebook.

Path resolution priority (used by `cw_load`, the editor, transformations):

1. Explicit `path =` argument
2. R option `brfssTools.config_path`
3. Environment variable `BRFSSTOOLS_CONFIG_PATH`
4. `tools::R_user_dir("brfssTools", "config")`
5. Package's `inst/extdata/` (read-only fallback)

## Crosswalk schema

`crosswalk.csv` — one row per (concept, year, state_var, cdc_var):

`concept_id, year, state_var, cdc_var, is_primary, source, score, notes`

- `concept_id` — the Platonic ideal name. Used as the column name in
  `brfss_pull()` output. Multiple state_vars across years can share a
  `concept_id` (e.g., `ACEADNED1/2/3/4` all → `ACEADNED`).
- `is_primary` — 1 for the canonical mapping per (year, state_var); 0
  for secondary one-to-many mappings.

Auxiliary files share the schema where applicable:

- `state_only.csv` — `concept_id, year, state_var, source, notes`
- `cdc_only.csv` — `concept_id, year, cdc_var, cdc_label, cdc_question,
  source, research_finding`
- `pending.csv` — undecided F_unmatched rows for later triage

## What's been built

### Done

| Component | Status |
|---|---|
| Pass 1 crosswalk (CDC raw vars ↔ OR vars, 2012-2024) | ✓ 1,872 rows |
| Concept-by-year pivot | ✓ 2,141 concepts |
| Taxonomy (12 domains × 55 subdomains × 73 categories + 3 tag systems) | ✓ |
| Static read-only HTML dashboard | ✓ |
| `concept_id` schema + auto-population on load | ✓ |
| Data-layer CRUD (cw_load, cw_save, cw_add_pair, etc.) | ✓ |
| Bulk concept rename (`cw_rename_concept`) | ✓ |
| Shiny editor — CDC-anchored heatmap + edit panel | ✓ |
| Shiny editor — State-anchored mode toggle | ✓ |
| Shiny editor — concept_id field + rename action | ✓ |
| Hybrid path resolution (user-dir + project-local override) | ✓ |
| `brfss_init_state()` for new users | ✓ |
| State codebook validator (`brfss_validate_state_codebook`) | ✓ |
| Transformation hooks: functional `.R` files | ✓ Stage 2 |
| Transformation hooks: declarative YAML categorical maps | ✓ Stage 3a |
| YAML → R code renderer for audit/docs | ✓ |

### Not yet done

| Component | Stage |
|---|---|
| Wire transformations (YAML + .R) into `brfss_pull()` | 3b |
| Categorical-map GUI tab in the editor | 4 |
| Pass 2 — CDC calculated vars crosswalk | backlog |
| Update `R/pull.R` to use `concept_id` | backlog |
| Race load-time calculator (now superseded by transformation hooks) | replaced |
| 36 pending F_unmatched rows triage | backlog |
| `year` → `vintage` rename in function args | backlog |
| Collapse `brfss_download()` into `brfss_pool()` | backlog |
| HAVHPAD 2012 → COPD pair verification | backlog |

## User-facing functions (current)

```{r}
# Initialization
brfss_init_state(state, state_codebook_path = NULL, path = NULL, overwrite = FALSE)
brfss_config_path(path = NULL, must_exist = TRUE)
sampled_states()
brfss_validate_state_codebook(df, strict = FALSE)

# Crosswalk CRUD (callable from console; also used by the editor)
cw_load(path = NULL)
cw_save(bundle, path = NULL)
cw_add_pair(bundle, year, state_var, cdc_var, concept_id = NULL, ...)
cw_remove_pair(bundle, year, state_var, cdc_var)
cw_update_pair(bundle, year, state_var, cdc_var, concept_id = NULL, ...)
cw_mark_state_only(bundle, year, state_var, concept_id = NULL, ...)
cw_mark_cdc_only(bundle, year, cdc_var, concept_id = NULL, ...)
cw_replace_cdc_partner(bundle, year, state_var, old_cdc_var, new_cdc_var, ...)
cw_replace_state_partner(bundle, year, old_state_var, cdc_var, new_state_var, ...)
cw_rename_concept(bundle, from, to)
cw_recompute_concept_ids(bundle)

# Editor
brfss_crosswalk_editor(path = NULL, launch.browser = TRUE, port = NULL)

# Transformations: functional (Stage 2)
brfss_setup_transformation(name, template = NULL, path = NULL, edit = TRUE)
brfss_setup_race(path = NULL, edit = TRUE)
brfss_list_transformations(path = NULL)

# Transformations: declarative YAML (Stage 3a)
brfss_setup_categorical_map(name, template = NULL, path = NULL, edit = TRUE)
brfss_setup_race_map(path = NULL, edit = TRUE)
brfss_load_categorical_map(fp)
brfss_render_transformation_code(name, path = NULL, save = FALSE)

# Original package functions (preserved; not yet updated for new concept_id schema)
brfss_pull(...)
brfss_download(...)
brfss_pool(...)
brfss_inventory_pool(...)
brfss_set_pool(...)
brfss_crosswalk(...)
brfss_crosswalk_audit(...)
```

## Decisions on file (so we don't relitigate)

- **One state, one crosswalk.** No `crosswalks/{state}.csv` directory. If
  multi-state ever happens, analysis functions take a `crosswalk =` arg.
- **CDC is universal anchor.** Always.
- **Variable names use `state_var` (generic), not `oha_var`.** Set up so a
  WA analyst could replace the file content without renaming columns.
- **`concept_id` is the unit of analysis.** R-friendly naming enforced.
  Defaults: paired/cdc_only → `cdc_var`, state_only → `state_var`. User
  edits override.
- **Storage: hybrid user-dir.** Default is `tools::R_user_dir`, overridable
  per-call with `path =`. Project-local for git-tracked work.
- **Transformations: both supported.** Functional `.R` files and
  declarative YAML categorical maps. `brfss_pull()` resolves YAML first.
- **YAML dependency: Suggests, not Imports.** Graceful error if missing.
- **Race: declarative YAML by default.** CDC-style spec, year-aware via
  `by_year:` blocks. The functional `.R` is the escape hatch.
- **Audit trail: git, not internal log.** No internal change log; the
  user's clone of the config dir is just files in git.
- **Backups: `.bak` rotation on every save** of crosswalk.csv.

## Known frictions / open questions

- `cw_save` writes to whatever path the bundle was loaded from. After
  `brfss_init_state`, that's the user dir. If you want to publish your
  Oregon work back to the package, you'd `cw_save(b, path = "...inst/extdata/")`
  with the explicit path — but the safety check refuses package paths,
  so we'd need a `package_dev = TRUE` flag for maintainers.
- Stage 3b will need careful surgery on `R/pull.R`. The function should
  detect when a requested name resolves to a transformation file vs. a
  crosswalk concept_id, and pull the transformation's input variables
  before applying it.
- The transformation hot-reload model (re-source on every pull) is fast
  during iteration but could be a perf issue on huge multi-year pulls.
  Probably fine; flag if it becomes a problem.

## How a new user gets started (the workflow we designed)

```{r}
install.packages("brfssTools")  # or pak::pak("nfparsons/brfssTools")
library(brfssTools)

# First-run init using the package's Oregon sample:
brfss_init_state("OR")

# Or for a fresh state with your own codebook:
brfss_init_state(
  state = "WA",
  state_codebook_path = "~/data/wa_brfss_codebook.csv"
)

# Open the editor:
brfss_crosswalk_editor()
# Map variables in the GUI, hit save.

# Set up a race transformation (CDC-style template):
brfss_setup_race_map()
# Edit the YAML, save.

# (After Stage 3b lands:)
df <- brfss_pull(
  vars  = c("ACEADNED", "race", "AGE"),
  years = 2019:2024
)
```

## Next session pickup

If this conversation ends and we resume in a new one, the resume point is:

> **Stage 3b: wire transformations into `brfss_pull()`.**
>
> Need to (1) read the existing `R/pull.R` to understand its current contract,
> (2) add resolution logic so a requested name like `"race"` is detected as
> a transformation rather than a crosswalk concept_id, (3) pull the
> transformation's input variables (parsing `inputs:` aliases for YAML or
> the `.R` file's input metadata for functional), (4) apply the transformation
> per-year after the data is loaded.
>
> Stage 4 (the categorical-map GUI tab) comes after.


```{r scratchpad}
devtools::document()
devtools::load_all()

brfss_list_transformations()
# Should show the race.yaml and race.R you have in your config dir
# (race.R as a separate row since it's a .R file in transformations/)
```

```{r}
fp <- file.path(brfss_config_path(), "transformations", "race.yaml")
spec <- brfss_load_categorical_map(fp)

test_data <- tibble::tibble(
  `_MRACE1`  = c(1, 2, 7, 99),
  `_HISPANC` = c(2, 2, 2, 1)
)
result <- brfssTools:::.brfss_apply_categorical_map(test_data, spec, year = 2023)
print(result)
```


