---
title: "Getting started with brfssTools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with brfssTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## What this package solves

The Behavioral Risk Factor Surveillance System (BRFSS) is the largest continuously-conducted health survey system in the world, but working with its longitudinal microdata is a slog. Variable names change between years, response codes shift (a `1` meaning "Yes" in 2015 might mean "No" in 2016 for the same concept), question wording drifts, and modules come and go. The transition from summary race columns to REAL-D arrays between 2021 and 2024 fractured demographic data into dozens of disparate columns. And if your work spans the public CDC national file *and* one or more state-restricted files, the bookkeeping multiplies.

`brfssTools` is a unified harmonization engine. It lets you treat BRFSS like `tidycensus` treats the ACS: register or download a pool of raw survey files, search a master crosswalk for the concepts you care about, and pull harmonized tibbles ready for analysis — across years, across datasets.

The package does **not** bundle BRFSS microdata. State files are restricted, and the national file is too large to ship with an R package. Instead, `brfssTools` operates on a **bring-your-own-data** (BYOD) model for state files, and provides a download helper for the public CDC national file.

## The three pieces

The package has three loosely-coupled pieces:

1. **The pool** — a directory of raw `.csv` (Oregon convention) or `.xpt` (CDC convention) files registered with `brfss_set_pool()`. One pool per dataset (`"National"`, `"OR"`, etc.).
2. **The master crosswalk** — a bundled set of CSVs (`concepts`, `concept_map`, `inventory`, `values`) loaded with `brfss_crosswalk()`. This is the Rosetta Stone that translates raw variables into stable concept IDs and applies recode rules.
3. **The pull engine** — `brfss_pull()` ties the two together: given concept IDs and a dataset, it fetches the right yearly files, applies recodes, and returns a tibble.

Most of what follows is showing how those three pieces compose.

## A fully self-contained example

So you can see the data shape and run code as you read, this vignette uses a small synthetic dataset rather than real BRFSS files. We'll create two years of OR-like CSVs in `tempdir()`, register them as a pool, build a tiny crosswalk in memory, and pull harmonized data.

```{r build-fake-pool}
library(brfssTools)
library(dplyr)
library(tibble)

# ---- a tiny synthetic OR pool ----
or_dir <- file.path(tempdir(), "vignette_or_pool")
dir.create(or_dir, showWarnings = FALSE)

writeLines(c(
  "SEQNO,ASTHNOW,GENHLTH,MENTHLTH",
  "2022000001,1,2,3",
  "2022000002,2,1,0",
  "2022000003,9,3,30",
  "2022000004,1,5,15"
), file.path(or_dir, "BRFSS_2022.csv"))

writeLines(c(
  "SEQNO,ASTHNOW,GENHLTH,MENTHLTH",
  "2023000001,2,1,0",
  "2023000002,1,4,8",
  "2023000003,1,2,2"
), file.path(or_dir, "BRFSS_2023.csv"))

brfss_clear_pool()                       # tidy any prior session state
brfss_set_pool("OR", or_dir)
brfss_pool_status()
```

Now a tiny crosswalk that knows about three concepts. In real use you'd call `brfss_crosswalk()` to load the bundled one — we're constructing it explicitly here so you can see the data structure.

```{r build-fake-cw}
cw <- list(
  inventory = tribble(
    ~source, ~year, ~raw_var_name, ~position, ~var_kind,  ~question_text,    ~module, ~missing_values_raw, ~source_file,    ~parse_notes,
    "core",  2022L, "ASTHNOW",     2L,        "categorical","Currently has asthma",NA, "9",                 "BRFSS_2022.csv", NA,
    "core",  2022L, "GENHLTH",     3L,        "categorical","General health",     NA, NA_character_,       "BRFSS_2022.csv", NA,
    "core",  2022L, "MENTHLTH",    4L,        "numeric",    "Mental health days", NA, NA_character_,       "BRFSS_2022.csv", NA,
    "core",  2023L, "ASTHNOW",     2L,        "categorical","Currently has asthma",NA, "9",                 "BRFSS_2023.csv", NA,
    "core",  2023L, "GENHLTH",     3L,        "categorical","General health",     NA, NA_character_,       "BRFSS_2023.csv", NA,
    "core",  2023L, "MENTHLTH",    4L,        "numeric",    "Mental health days", NA, NA_character_,       "BRFSS_2023.csv", NA
  ),
  values = tribble(
    ~source, ~year, ~raw_var_name, ~code, ~label,    ~is_missing,
    "core",  2022L, "ASTHNOW",     "9",   "Refused", TRUE,
    "core",  2023L, "ASTHNOW",     "9",   "Refused", TRUE
  ),
  concepts = tribble(
    ~concept_id, ~concept_label,         ~domain,   ~subdomain, ~construct_notes, ~canonical_response_type,
    "ASTHNOW",   "Currently has asthma", "health",  NA,         NA,               "yes_no",
    "GENHLTH",   "Self-rated health",    "health",  NA,         NA,               "ordinal",
    "MENTHLTH",  "Poor mental health days (last 30)", "mental_health", NA, NA,    "count"
  ),
  concept_values = tibble::tibble(
    concept_id = character(), code = character(), label = character()
  ),
  concept_map = tribble(
    ~concept_id, ~source, ~year, ~raw_var_name, ~question_text_drift, ~recode_type, ~recode_rule, ~notes,
    "ASTHNOW",   "core",  2022L, "ASTHNOW",     "identical",          "inline",     "1=1; 2=0",   NA,
    "ASTHNOW",   "core",  2023L, "ASTHNOW",     "identical",          "inline",     "1=1; 2=0",   NA,
    "GENHLTH",   "core",  2022L, "GENHLTH",     "identical",          "identity",   NA,           NA,
    "GENHLTH",   "core",  2023L, "GENHLTH",     "identical",          "identity",   NA,           NA,
    "MENTHLTH",  "core",  2022L, "MENTHLTH",    "identical",          "identity",   NA,           NA,
    "MENTHLTH",  "core",  2023L, "MENTHLTH",    "identical",          "identity",   NA,           NA
  ),
  lake = "brfss"
)
```

With the pool registered and the crosswalk in hand, a pull is one call:

```{r pull-wide}
brfss_pull(
  cw,
  concept_ids = c("ASTHNOW", "GENHLTH", "MENTHLTH"),
  dataset     = "OR",
  id_cols     = "SEQNO",
  output      = "wide"
)
```

A few things happened in that call worth noticing:

- The 2022 `ASTHNOW = 9` row came back as `NA` because we tagged code `9` as `is_missing = TRUE` in the values table.
- The 2022 `ASTHNOW = 1` rows came back as `"1"` and `ASTHNOW = 2` as `"0"`, because the inline recode rule mapped `1 -> 1` and `2 -> 0` (Yes/No standardization).
- `GENHLTH` and `MENTHLTH` came through unchanged because their recode type was `"identity"`.

You can also pull in long format, which is more convenient for some downstream uses:

```{r pull-long}
brfss_pull(
  cw,
  concept_ids = c("ASTHNOW", "GENHLTH"),
  dataset     = "OR",
  id_cols     = "SEQNO",
  output      = "long"
) |>
  head(10)
```

## Real-world setup

The runnable example above shows the mechanics. In practice you don't construct the crosswalk yourself — you load the bundled one — and your pool points at real files. There are two pathways.

### National CDC data

The public combined landline + cell phone LLCP file is downloadable. `brfss_download()` fetches the requested years from CDC, extracts them into your user cache, and registers the cache as the `"National"` pool in one step. You need the `haven` package installed (it ships as a Suggest) because the files are SAS XPT.

```{r national-download, eval = FALSE}
# One-time setup, per machine, per year
brfss_download(2020:2024)
#> Downloading BRFSS 2020 from CDC...
#>   -> ~/.cache/R/brfssTools/LLCP2020.XPT
#> Downloading BRFSS 2021 from CDC...
#>   -> ~/.cache/R/brfssTools/LLCP2021.XPT
#> ...
#> Indexed 5 file(s) for dataset 'National'.

# Now usable like any other dataset
cw <- brfss_crosswalk(dataset = "National", year = 2020:2024)

national_data <- brfss_pull(
  cw,
  concept_ids = c("survey_weight", "psu", "strata", "GENHLTH", "MENTHLTH"),
  dataset     = "National",
  states      = c("OR", "WA", "ID", "CA"),   # USPS or FIPS, your choice
  id_cols     = "SEQNO",
  output      = "wide"
)
```

Years before 2011 are not supported — the 2011 weighting methodology change makes earlier files non-comparable. The cache lives at `brfss_cache_dir()`, which is the OS-appropriate path returned by `tools::R_user_dir()`.

### State-restricted data (BYOD)

For Oregon and other state-restricted files, point the package at your secure data directory. Files can be `.csv` or `.xpt`; year is auto-detected from the filename or the `SEQNO` column.

```{r state-byod, eval = FALSE}
brfss_set_pool("OR", "Z:/Secure_Data/BRFSS/Raw_Files")

cw <- brfss_crosswalk(dataset = "OR", year = 2018:2024)

oregon_data <- brfss_pull(
  cw,
  concept_ids = c("survey_weight", "psu", "strata",
                  "multnomah_flag", "fmd", "GENHLTH"),
  dataset     = "OR",
  id_cols     = "SEQNO",
  output      = "wide"
)
```

You can have multiple pools registered at once — `brfss_pool_status()` will show them all.

## Inspecting a crosswalk: search and lookup

Before showing the bundled real-world crosswalk, let's exercise the inspection helpers against the synthetic `cw` we built above — same API, just less data. `brfss_search()` does case-insensitive regex matching against concept labels, raw variable names, and question text:

```{r search-demo}
brfss_search(cw, "asthma")
brfss_search(cw, "^MENT", scope = "raw_var")
```

`brfss_lookup()` returns the full crosswalk record(s) for one or more concept IDs, joined with concept and inventory metadata — handy when you need to know which raw variable backs a concept in each year and what recode rule applies:

```{r lookup-demo}
brfss_lookup(cw, "ASTHNOW") |>
  select(concept_id, source, year, raw_var_name, recode_type, recode_rule)
```

## Inspecting the bundled master crosswalk

In real use, you load the bundled crosswalk rather than hand-building one. The same helpers operate on it:

```{r search-and-lookup, eval = FALSE}
cw <- brfss_crosswalk()

# Search by free text against concept labels, raw variable names, or
# question text. Case-insensitive, regex-aware.
brfss_search(cw, "asthma")
brfss_search(cw, "^MENT", scope = "raw_var")
brfss_search(cw, "depression|anxiety", scope = "concept")

# Pull the full crosswalk record(s) for a concept
brfss_lookup(cw, "GENHLTH")
```

## The `source` column: how datasets coexist

Every row in the concept_map carries a `source` tag controlling where the rule applies:

| `source` value      | Meaning                                                        |
|---------------------|----------------------------------------------------------------|
| `"core"`            | Applies to every dataset (national LLCP and any state)         |
| `"OR"`              | Oregon state-added items only                                  |
| `"WA"`, `"CA"`, ... | State-added items for that state (when added to the registry) |

When you call `brfss_pull(..., dataset = "X")`, the engine filters the concept_map to rows where `source %in% c("core", "X")`. So:

- A National pull sees `core` (plus any rules tagged `"National"`, e.g., CDC-only calculated variables).
- An Oregon pull sees `core` rules *plus* OR-specific items like state-added modules or local geography flags.

This is what makes the architecture extensible. Adding support for a new state means tagging that state's added rules with the state code and registering the dataset in `brfss_datasets()` — no changes to any pull/search/lookup code.

```{r datasets-listing}
brfss_datasets()
```

## Pulling concepts: common patterns

The example above already showed wide vs long output and missing-code handling. A few more patterns worth knowing.

### Year filtering

If your concept_map covers many years but you only want a subset, pass `years`:

```{r year-filter}
brfss_pull(
  cw,
  concept_ids = "GENHLTH",
  dataset     = "OR",
  year        = 2023,           # only this year
  id_cols     = "SEQNO"
)
```

This filters at the concept_map level *before* fetching files, so it's an efficiency win when you have a many-year pool but a single-year analysis.

### State filtering (national pulls)

Most useful with the national file, where the `_STATE` column carries FIPS codes. You can pass either USPS abbreviations or FIPS integers:

```{r states-syntax, eval = FALSE}
# Equivalent
brfss_pull(cw, "GENHLTH", dataset = "National", states = c("OR", "WA"))
brfss_pull(cw, "GENHLTH", dataset = "National", states = c(41, 53))
```

State filtering is silently ignored if the file has no `_STATE` column.

### Keeping raw values alongside harmonized

Useful for auditing recode rules:

```{r keep-raw}
brfss_pull(
  cw,
  concept_ids = "ASTHNOW",
  dataset     = "OR",
  id_cols     = "SEQNO",
  keep_raw    = TRUE,
  output      = "long"
) |>
  head()
```

You can see the raw `1`/`2`/`9` values next to their harmonized `1`/`0`/`NA`.

## Demographic cleaners

The `brfss_clean_*` family handles the most drift-prone demographic variables. They all share a design: take the wide tibble produced by `brfss_pull()`, look for known concept-id columns, append a standardized derived column.

### Race and ethnicity

OR's analyst-facing data uses the BRFSS **MRACE code system** (10 White, 20 Black, 30 AI/AN, 40 Asian umbrella, 41–47 Asian subgroups, 50 PI umbrella, 51–54 PI subgroups, 60 Other). Across years the *naming* shifts but the codes stay consistent: `RACE1`–`RACE10` (pre-2017), `MRACE1_1`–`MRACE1_14` (2017–2021), `MRACE2_01`–`MRACE2_14` (2022–2024). The crosswalk maps all of them under the canonical `race_array_*` concepts so callers don't need to know the year-specific naming.

Hispanic origin lives in a parallel array: a single `HISPANC3` column (2013–2014) or four sub-flag columns (`HISPANC3_1`–`HISPANC3_4` for 2017–2021, `HISPANC3_01`–`HISPANC3_04` for 2022–2024), all carrying HISPANC3 codes (1 Mexican, 2 Puerto Rican, 3 Cuban, 4 Other Hispanic, 5 No, 7 DK, 9 Refused). These are mapped under canonical `hisp_array_*` concepts.

`brfss_clean_race()` scans whatever `race_array_*` and `hisp_array_*` columns the wide tibble contains, identifies Hispanic membership from any `hisp_array_*` slot showing 1–4, and applies one of three standards:

```{r race-derivation, eval = FALSE}
demos <- brfss_pull(
  cw,
  concept_ids = c(paste0("race_array_", 1:14),
                  paste0("hisp_array_", 1:4)),
  dataset = "OR",
  output  = "wide"
)

# Three standards, choose based on reporting need:
demos |> brfss_clean_race(standard = "cdc")     # 8-level CDC mutually-exclusive
demos |> brfss_clean_race(standard = "fewest")  # White NH vs BIPOC vs Hispanic
demos |> brfss_clean_race(standard = "reald")   # Asian / PI subgroup indicators
```

The three standards each produce different output columns: `cdc` adds `race_eth_cdc`, `fewest` adds `race_eth_fewest`, and `reald` adds a set of `reald_*` binary indicators for each Asian and Pacific Islander subgroup separable in MRACE encoding (`reald_chinese`, `reald_filipino`, `reald_native_hawaiian`, `reald_samoan`, etc.).

**Granularity caveat.** MRACE is a CDC-collapsed encoding. Sub-detail that exists in OR's REX-level questionnaire — Slavic, Eastern European, Afro-Caribbean, Ethiopian, Somali, Hmong, Marshallese, Middle Eastern, North African — is rolled up before it lands in the analyst-facing file (e.g., MENA codes 80–82 collapse into MRACE 10 White). That detail is not recoverable from these columns. If you need full REAL-D detail, you'd need raw REX1–REX10 added to the data file, which is a data-shop request, not a package fix.

### Age groupings

BRFSS exposes age in several drift-prone forms across years (continuous top-coded `_AGE80`, 13-level `_AGEG5YR`, six-level `_AGE_G`, and a 65+ binary `_AGE65YR`). `brfss_clean_age()` picks from a menu of common groupings and uses whichever source column is available for each row, in a waterfall:

```{r age-derivation, eval = FALSE}
demos <- brfss_pull(
  cw,
  concept_ids = c("age_continuous", "age_5yr", "age_6group"),
  dataset = "OR", output = "wide"
)

demos |> brfss_clean_age(scheme = "5yr")        # 13 categories
demos |> brfss_clean_age(scheme = "6group")     # CDC default 6 categories
demos |> brfss_clean_age(scheme = "65plus")     # binary
demos |> brfss_clean_age(scheme = "samhsa")     # NSDUH/SAMHSA convention
demos |> brfss_clean_age(scheme = "suicide")    # suicide-surveillance convention

# Custom breaks
demos |> brfss_clean_age(
  scheme = "custom",
  breaks = c(18, 30, 50, 70, Inf),
  labels = c("18-29", "30-49", "50-69", "70+"),
  output_col = "age_4cat"
)

# Multiple groupings on the same dataset
demos |>
  brfss_clean_age(scheme = "5yr",    output_col = "age_5yr_grp") |>
  brfss_clean_age(scheme = "65plus", output_col = "age_65")
```

Each row uses the best available source. If `age_continuous` is present and non-NA, it's used; otherwise the function falls through to `age_5yr`, then `age_6group`, then `age_65plus`. Schemes that need bracket boundaries the categorical sources don't expose (`samhsa`, `custom`) require continuous age and error early if it's missing.

## Onboarding a new dataset

This is one of the highest-leverage pieces. Suppose someone hands you Washington State BRFSS files. You want to add `"WA"` as a first-class dataset.

### Step 1: register the dataset

Edit `R/datasets.R` and add a row to the registry:

```{r registry-edit, eval = FALSE}
# In R/datasets.R, .brfss_dataset_registry()
"WA", "Washington State BRFSS (restricted; BYOD)", "csv", FALSE, FALSE
```

After reinstalling, `brfss_set_pool("WA", path)` will not warn about an unknown dataset and `brfss_pull(dataset = "WA")` will work.

### Step 2: register the pool and scan it

```{r inventory-walkthrough, eval = FALSE}
brfss_set_pool("WA", "/secure/wa_brfss")

# Generate a starting inventory: all variables, by year, with positions
# and inferred types. Same shape as raw_inventory_brfss.csv.
wa_inv <- brfss_inventory_pool("WA")

# Scan observed values + frequencies for select variables. Same shape
# as raw_values_brfss.csv. (Slow on large national files; restrict to
# vars of interest with `vars =`.)
wa_vals <- brfss_values_pool("WA",
                             vars = c("ASTHNOW", "_AGEG5YR", "WA_ADDED_1"))
```

### Step 3: diff against the existing crosswalk

Find variables in the WA pool that aren't yet in the master crosswalk:

```{r diff-crosswalk, eval = FALSE}
cw <- brfss_crosswalk()

new_to_wa <- dplyr::anti_join(
  wa_inv,
  cw$inventory,
  by = c("year", "raw_var_name")
)
```

Anything in `new_to_wa` is a candidate for a `source = "WA"` rule (state-added). Anything *already* in `cw$inventory` is presumably shared LLCP and needs no work — those rules already apply to WA via the `core` tag.

### Step 4: write the crosswalk additions

For each WA-added variable, add rows to `concept_map_brfss.csv` (and any new concepts to `concepts_brfss.csv`, value labels to `concept_values_brfss.csv`, etc.) with `source = "WA"`. Reinstall, and `brfss_pull(dataset = "WA")` will start applying them.

The point of this whole architecture: adding a state is a data-curation task, not a code-modification task.

## Pool management

Multi-pool sessions get a real status view:

```{r pool-management}
brfss_pool_status()
```

Adding more pools (here, a synthetic National pool):

```{r add-second-pool}
nat_dir <- file.path(tempdir(), "vignette_nat_pool")
dir.create(nat_dir, showWarnings = FALSE)
writeLines(c("SEQNO,_STATE,GENHLTH",
             "2022000001,41,2",
             "2022000002,53,3",
             "2022000003,6,1"),
           file.path(nat_dir, "LLCP2022.csv"))

brfss_set_pool("National", nat_dir)
brfss_pool_status()
```

Clearing one or all pools:

```{r clear-pool}
brfss_clear_pool("National")
brfss_pool_status()

brfss_clear_pool()           # nuke them all
brfss_pool_status()
```

## Where to go from here

`brfss_pull()` automatically pulls survey-design variables alongside the concepts you ask for: as long as `survey_weight`, `psu`, and `strata` exist in the crosswalk, they're added to your wide-format result. (For OR data the strata column is missing for most years — `brfss_as_svy()` handles that case below.) `SEQNO` is included by default via `id_cols`. So a typical pull is already analysis-ready:

```{r srvyr-handoff, eval = FALSE}
library(srvyr)

or_data <- brfss_pull(
  cw, c("ASTHNOW", "GENHLTH"),
  dataset = "OR", output = "wide"
  # SEQNO + survey_weight + psu (+ strata if present) are auto-included
)

or_svy <- brfss_as_svy(or_data)

or_svy |>
  filter(!is.na(GENHLTH)) |>
  group_by(GENHLTH) |>
  summarize(asthma_prev = survey_mean(ASTHNOW == "1",
                                      na.rm   = TRUE,
                                      vartype = "ci"))
```

`brfss_as_svy()` handles two real-world quirks: weights come back as character (the recode engine works on strings) and get coerced to numeric; OR's missing-strata case is detected and the design built without stratification.

For multi-year trend analyses you'll typically want to bind several years together — `brfss_pull()` does this for you when you pass a multi-year range, but you can also pull per-year and stack manually if you need year-specific weights or design adjustments.

The bundled crosswalk is a living artifact. As BRFSS evolves and as you tag more state-added rules, the map grows. Contributing back to it (or shipping your shop's local fork) is how the package gets more useful over time.
